Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-11-19 analysis/paper/paper.Rmd
minor edits to bioenergetics

An estimated 200 million juvenile salmonids emigrate through the lower Snake and Columbia rivers, annually. Of those, approximately 16.4 million (8%) are consumed by northern pikeminnow [Beamesderfer1996]. @Petersen1994 estimated that the annual loss of juvenile salmonids to northern pikeminnow to be 7.3% of all juvenile salmonids entering the John Day Reservoir. Additionally, 78% of these salmonids consumed near a dam were consumed when alive [@Petersen1994]. Dams are known to be areas of predation, but @Ward1995 estimated that 48% of predation is occurring away from dams in mid-reservoir areas. Most of the research has been targeted around dams and reservoirs, but there are areas functioning similar to a reservoir in the rivers upstream. One of these areas is the Deadwater Slough in the Salmon River. The Deadwater Slough is a 25.3-acre area that is unnaturally slow and deep within the Salmon River and resembles a small reservoir (Figure \@ref(fig:map)). This unnatural area has little to no cover, higher depth, and fine substrate which favor piscivorous fish predators (e.g., northern pikeminnow *Ptychocheilus oregonensis*, smallmouth bass *Micropterus dolomieu*) [@Watkins2015]. @Axel2015 demonstrated decreased rates of emigration and apparent survival for Sockeye Salmon *Oncorhynchus nerka* from Redfish Lake emigrating during the spring. Further, recent winter telemetry studies have indicated decreased transition probabilities (approximately 10% less than surrounding reaches) of juvenile Chinook salmon through Deadwater Slough during fall and winter months [@Ackerman2018; @Porter2019].

@@

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M^2_i}{\sum\limits_{i = 1}^k m_i M_i}
$$
We additionally calculated the proportional stock density (PSD) of northern pikeminnow with 300 mm total length (TL) for stock and 400 mm TL for quality.
$$
PSD_{i} = 100 * \frac{FQ_{i}}{FS_{i}}
$$
where $FQ_{i}$ is the number of fish $\ge$ quality-length for species $i$, and $FS_{i}$ is the number of fish $\ge$ stock-length for species $i$.

@@

To estimate the impacts from northern pikeminnow on juvenile Chinook Salmon outmigrants in Deadwater Slough, we used an R-based application of Fish Bioenergetics v4.0 developed by @Deslauriers2017. We estimated a daily rate of consumption for an individual northern pikeminnow based on predator and prey energy densities, water temperatures, and predator start and end weights. We also ran two alternative models one with a year of temperature inputs from 2019, and another with the same temperature inputs but a 10% growth in weight of the average pikeminnow at the end of the model. The average length of northern pikeminnow captured at the Deadwater Slough during our study was 370.6 mm which calculates to an average starting weight of 847 g using the FSA package (@Ogle2021) in R which provides a weight-length formula for northern pikeminnow sourced from @Parker1995. Predator energy density for northern pikeminnow was available in the Bioenergetics v4.0 application and was fixed at 6,703 Joules(J)/g. Prey energy densities was taken from @Moss2016 where they estimated juvenile Chinook Salmon at 21,500 J/g. Using these temperatures, weights, and energy inputs, we estimated a total consumption in grams for an individual norther pikeminnow within Deadwater Slough.

We ran three bioenergetic models for this study. First, we chose a 78-day period from September 1 through November 17 where we know Chinook Salmon DSR emigrants begin to enter the mainstem Salmon River from natal tributaries (e.g., Lemhi River) and begin to migrate downstream, but when temperatures are still high enough that juveniles are not yet exhibiting concealment behavior or torpor. The second model was still assuming no growth but taking the entire year into account for the model rather than the 78-day fall period. The assumption of no growth for pikeminnow was included to show what may be occurring if the population of northern pikeminnow is stable. If growther of the population is occurring, the alternative model shows how an increase of 10% body weight changes the estimates of consumption over a full year period. To estimate the number of possible Chinook Salmon that could be consumed, we assumed a weight of 12 g per individual chinook Salmon. This weight was an average from juvenile Chinook salmon emigrating past six rotary screw traps operating in the Upper Salmon above Deadwater Slough including traps in the Lemhi, Pahsimeroi, and North Fork Salmon rivers and one trap operating near the Sawtooth hatchery.

@@

We were able to gastric lavage the majority of fish collected including some non-predatory species. We euthanized nine northern pikeminnow for dissection to confirm that the gastric lavage was effective in removing stomach contents. Our results support the findings the previous year that dissected individuals that the dissected gastric lavage was effective in removing stomach contents (@Lott2020). Additionally , we found that the fish captured in trap-nets had a similar proportion of stomach content samples. Of those, 1,214 (76.2%) were completely empty, 345 (21.6%) had stomach contents, and 35 (2.2%) contained fish parts. We were able to identify juvenile shiners, a largescale sucker, sculpin, mountain whitefish, and one Chinook Salmon within stomach contents, but most samples were too digested to identify to species.

@@

Although not formally assessed in this study, avian predators including Great Blue Herons *Ardea herodias* and Bald Eagles *Haliaeetus leucocephalus* are another potential source of mortality for juvenile salmon in the Deadwater Slough. The Deadwater Slough is recognized as an important bird watching and nesting area due to the riparian and backwater habitats created by the feature [(Deadwater Slough - Audubon Important Bird Areas)](https://www.audubon.org/important-bird-areas/deadwater-slough#). Several piscivorous bird species have been documented using the Deadwater Slough that include the Common *Mergus merganser* and Hooded *Lophodytes cucullatus* mergansers, the Great Blue Heron, the Double-crested Cormorant *Phalacrocorax auritus*, and the Belted Kingfisher *Megaceryle alcyon* [@eBird2021]. During the intial sampling event in 2019, a two-person crew walked the entire reach and surrounding and upstream areas scanning for passive integrated transponder (PIT) tags. During that informal survey, nine PIT tags were recovered near active bird nests or in an upstream anastomizing reach where herons and eagles are prevalent, suggesting that mortality may have been a result of avian predation. The PIT tag histories in [PTAGIS](https://www.ptagis.org/) indicate these tags were implanted into a combination of juvenile Chinook Salmon (3), Sockeye Salmon (3), and steelhead (3). Avian predation contributes a major component of the total mortality for yearling Chinook Salmon in some locations in the lower Snake River and Columbia River, particularly at hydroelectric dams and within reservoirs [@Evans2012; -@Evans2016]; however, we did not observe large colonies of piscivorous birds within the study area. Although there is documentation of individual Double-crested Cormorants [@eBird2021] at the Deadwater Slough, the site is not within their breeding range, rather, it is part of a migration corridor. Given the current avian species known to occupy Deadwater Slough, it is unlikely that avian predation on juvenile salmonids is comparable to elsewhere in the Columbia River basin with large piscivorous bird colonies. Nevertheless, we hypothesize that the reservoir-like conditions at the Deadwater Slough may increase the probability of avian predation on juvenile Chinook Salmon from the many piscivorous birds known to use the site. Future estimates of predation would benefit from consideration of the contribution of piscivorous avian predators.

@@

Is there anything in the literature that discusses habitat preferences for pike minnow? If we speed up velocities and add some cover potentially pike minnow predation success will be lowered. Will these fish just move elsewhere?

>>>>

Northern Pikeminnow are estimated to consume 8% (16.4 million) of the estimated 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually [@Beamesderfer1996]. Dams and reservoirs in the Columbia River are associated with high rates of predation on salmonids [@Petersen1994; @Ward1995]. Because migration rates of juvenile Chinook Salmon are reduced during reservoir passage [@Venditti2000], migrating smolts may become more vulnerable to predation in areas with slower velocities. Most of the research has been targeted around dams and reservoirs, but there are areas functioning similar to a reservoir in the ruvers upstream. One of these areas is the Deadwater Slough in the Salmon River. The Deadwater Slough is a 25.3-acre area that is unnaturally slow and deep within the Salmon River and resembles a small reservoir (Figure \@ref(fig:map)). This unnatural area has little to no cover, higher depth, and fine substrate which favor piscivorous fish predators (e.g., northern pikeminnow *Ptychocheilus oregonensis*, smallmouth bass *Micropterus dolomieu*) [@Watkins2015]. @Axel2015 demonstrated decreased rates of emigration and apparent survival for Sockeye Salmon *Oncorhynchus nerka* from Redfish Lake emigrating during the spring. Further, recent winter telemetry studies have indicated decreased transition probabilities (approximately 10% less than surrounding reaches) of juvenile Chinook salmon through Deadwater Slough during fall and winter months [@Ackerman2018; @Porter2019].

@@

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M^2_i}{\sum\limits_{i = 1}^k m_i M_i}
$$
We additionally calculated the proportional stock density (PSD) of Northern Pikeminnow with 300 mm total length (TL) for stock and 400 mm TL for quality.
$$
PSD_{i} = 100 * \frac{FQ_{i}}{FS_{i}}
$$
where $FQ_{i}$ is the number of fish $\ge$ quality-length for species $i$, and $FS_{i}$ is the number of fish $\ge$ stock-length for species $i$.

@@

To estimate the consumption of juvenile Chinook Salmon outmigrants by Northern Pikeminnow in the Deadwater Slough, we used the Fish Bioenergetics v4.0 applicatioon developed by @Deslauriers2017. The daily rate of consumption in grams for an individual Northern Pikeminnow was based on predator and prey energy densities, predator start and end weights, and water temperatures. Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 21,500 J/g for juvenile Chinook salmon based on estimate from @Moss2016. To calculate the average predator start weight, we converted the average length of Northern Pikeminnow captured during our study to a weight using a weight-length formula for Northern Pikeminnow [@Parker1995] from the FSA package [@Ogle2021]. Continuous water temperature data from 2019 were used in two alternative models, on using the baseline start and end weights and the second with a 10% increase in average end weight of Northern Pikeminnow.

We ran three bioenergetic models for this study. For the first model, we chose a 78-day period from September 1 through November 17 when DSR emigrants are known to enter the mainstem Salmon River from natal tributaries (e.g., Lemhi River) and begin their downstream migration. During this time, water temperatures exceed the range (0-7$^\circ$C) that would illicit concealment behavior or torpor from juvenile Chinook Salmon. This model assumed no growth for Northern Pikeminnow. The second model was run for one-year and assumed no growth for Northern Pikeminnow. The assumption of no growth for pikeminnow was included to show what may be occurring if the populaiton of Northern Pikeminnow is stable. If growth of the population is occurring, the alternative model shows how an increase of 10% body weight changes the estimtes of consumption over a full year period. To estimate the number of possible Chinook Salmon that could be consumed, we assumed a weight of 12 g per individual Chinook Salmon. This weight was an average from juvenile Chinook Salmon emigrating past six rotary screw traps operating in the Upper Salmon above Deadwater Slough including traps in the Lemhi, Pahsimeroi, and North Fork Salmon rivers and one trap operating near the Sawtooth hatchery.

<!--
To estimate the impacts from northern pikeminnow on juvenile Chinook Salmon outmigrants in Deadwater Slough, we used an R-based application of Fish Bioenergetics v4.0 developed by @Deslauriers2017. We estimated a daily rate of consumption for an individual northern pikeminnow based on predator and prey energy densities, water temperatures, and predator start and end weights. We also ran two alternative models one with a year of temperature inputs from 2019, and another with the same temperature inputs but a 10% growth in weight of the average pikeminnow at the end of the model. The average length of northern pikeminnow captured at the Deadwater Slough during our study was 370.6 mm which calculates to an average starting weight of 847 g using the FSA package (@Ogle2021) in R which provides a weight-length formula for northern pikeminnow sourced from @Parker1995. Predator energy density for northern pikeminnow was available in the Bioenergetics v4.0 application and was fixed at 6,703 Joules(J)/g. Prey energy densities was taken from @Moss2016 where they estimated juvenile Chinook Salmon at 21,500 J/g. Using these temperatures, weights, and energy inputs, we estimated a total consumption in grams for an individual norther pikeminnow within Deadwater Slough.

We ran three bioenergetic models for this study. First, we chose a 78-day period from September 1 through November 17 where we know Chinook Salmon DSR emigrants begin to enter the mainstem Salmon River from natal tributaries (e.g., Lemhi River) and begin to migrate downstream, but when temperatures are still high enough that juveniles are not yet exhibiting concealment behavior or torpor. The second model was still assuming no growth but taking the entire year into account for the model rather than the 78-day fall period. The assumption of no growth for pikeminnow was included to show what may be occurring if the population of northern pikeminnow is stable. If growther of the population is occurring, the alternative model shows how an increase of 10% body weight changes the estimates of consumption over a full year period. To estimate the number of possible Chinook Salmon that could be consumed, we assumed a weight of 12 g per individual chinook Salmon. This weight was an average from juvenile Chinook salmon emigrating past six rotary screw traps operating in the Upper Salmon above Deadwater Slough including traps in the Lemhi, Pahsimeroi, and North Fork Salmon rivers and one trap operating near the Sawtooth hatchery.
-->

@@

The majority of fish collected during this study received gastric lavage, including some non-predatory species. To validate the efficacy of this method, we euthanized nine Northern Pikeminnow after gastric lavage was completed and removed the remaining stomach contents via dissection. Our results support previous findings that gastric lavage effectively removes stomach contents (@Lott2020). Additionally, we found that the fish captured in net-traps had a similar proportion of stomach content samples. of those, 1,214 (76.2%) were completely emply, 345 (21.6%) had stomach contents, and 35 (2.2%) contained fish remnants. We were able to identify juvenile Red Shiners, a Largescale Sucker, Sculpin, Mountain Whitefish, and one Chinook Salmon within stomach contents, but most samples were too digested to identify to species.

@@

Although not formally assessed in this study, avian predators including Great Blue Herons *Ardea herodias* and Bald Eagles *Haliaeetus leucocephalus* are another potential source of mortality for juvenile salmon in the Deadwater Slough. The Deadwater Slough is recognized as an important bird watching and nesting area due to the associated riparian and backwater habitats [(Deadwater Slough - Audubon Important Bird Areas)](https://www.audubon.org/important-bird-areas/deadwater-slough#). Several piscivorous bird species have been documented using the Deadwater Slough that include the Common *Mergus merganser* and Hooded *Lophodytes cucullatus* mergansers, the Great Blue Heron, the Double-crested Cormorant *Phalacrocorax auritus*, and the Belted Kingfisher *Megaceryle alcyon* [@eBird2021]. During the intial sampling event in 2019, a two-person crew walked the entire reach and surrounding and upstream areas scanning for passive integrated transponder (PIT) tags. During that informal survey, nine PIT tags were recovered near active bird nests or in an upstream anastomizing reach where herons and eagles are prevalent, suggesting that mortality may have been a result of avian predation. The PIT tag histories in [PTAGIS](https://www.ptagis.org/) indicate these tags were implanted into a combination of juvenile Chinook Salmon (3), Sockeye Salmon (3), and steelhead (3). Avian predation contributes a major component of the total mortality for yearling Chinook Salmon in some locations in the lower Snake River and Columbia River, particularly at hydroelectric dams and within reservoirs [@Evans2012; -@Evans2016]; however, we did not observe large colonies of piscivorous birds within the study area. Although there is documentation of individual Double-crested Cormorants [@eBird2021] at the Deadwater Slough, the site is not within their breeding range, rather, it is part of a migration corridor. Given the current avian species known to occupy Deadwater Slough, it is unlikely that avian predation on juvenile salmonids is comparable to elsewhere in the Columbia River basin with large piscivorous bird colonies. Nevertheless, we hypothesize that the reservoir-like conditions at the Deadwater Slough may increase the probability of avian predation on juvenile Chinook Salmon from the many piscivorous birds known to use the site. Future estimates of predation would benefit from consideration of the contribution of piscivorous avian predators.

@@

<!--
Is there anything in the literature that discusses habitat preferences for pike minnow? If we speed up velocities and add some cover potentially pike minnow predation success will be lowered. Will these fish just move elsewhere? 
-->
